Implementing Reactive Systems Directly\\with I/O Automata in ioa++

### 2012-02-24 papers/paper2/programming_model.tex
Down to 17 pages.

In this section, we show how I/O automata can be used to construct component based systems using ioa++.
Section~\ref{representation} describes how I/O automata are represented in the C++ language.
We then discuss the mechanisms available for dynamic composition in Section~\ref{dynamic_composition}.
Section~\ref{constellations} concludes with techniques for managing dynamic constellations of automata.

\subsection{I/O Automata Representation\label{representation}}

As was mentioned in Section~\ref{design}, all user automata inherit from the common base class (ioa::automaton) which implements the system action interface.
\ifjournal
The following code declares the asynch\_lcr\_automaton:
\begin{lstlisting}
template <typename UID>
class asynch_lcr_automaton :
  public ioa::automaton
{
  ...
};
\end{lstlisting}
The asynch\_lcr\_automaton is declared as a C++ template where the UID template parameter must model the \emph{unique identifier} concept\footnote{The term ``concept'' is used here in the sense of generic programming in C++, i.e., a collection of type requirements imposed by a C++ template.}.
The generic programming facilities of the C++ language aid in the development of general purpose reusable components.

@@

\paragraph*{Declaring/defining actions.}
Actions are member variables that model the \emph{action} 
\ifjournal
concept.
\else
concept\footnote{The term ``concept'' is used here in the sense of generic programming in C++, i.e., a collection of type requirements imposed by a C++ template.}.
\fi
Actions contain methods for preconditions, effects, and scheduling along with \emph{traits} that indicate the action's type (input, output, internal), value status, value type, parameter status, and parameter type.
These traits are used to guide the execution of the methods by the dispatcher and to supply values and parameters as necessary.
The traits are also used to check that input actions are bound to output actions with the same value status and type.
Bindings are checked for type compatibility at compile time using the generic programming facilities of C++.

A number of wrappers and macros are provided by ioa++ for declaring actions.
%%To facilitate declaring actions, a number of wrappers and macros corresponding to the action types listed in Section~\ref{practical} are available.
Local actions are declared using three functions corresponding respectively to precondition, effect, and scheduling.
%%The macros assume the functions are named action\_name\_precondition, action\_name\_effect, and action\_name\_schedule.
\ifjournal
For example, the send action of the asynch\_lcr\_automaton is declared:
\begin{lstlisting}
private:
  bool send_precondition () const { ... }
  UID send_effect () { ... }
  void send_schedule () const { ... }
public:
  V_UP_OUTPUT (asynch_lcr_automaton, send, UID);
\end{lstlisting}
\fi
Input actions are declared similarly except for the absence of a precondition.
The precondition and scheduling functions are not allowed to change the state of the automaton and are consequently declared const.
The macros for actions that are subject to binding by other automata are declared public.

\paragraph*{Scheduling.}
Automata submit actions to the scheduler using the ioa::schedule function.
The ioa::schedule function can be called at any time.
However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
A common idiom is for each scheduling function in an automaton to call a common member function that schedules all enabled local actions to avoid replicating scheduling logic.
The common scheduling function consists of calls to ioa::schedule guarded by preconditions as described in Section~\ref{scheduling}.

@@

\subsection{Dynamics\label{dynamic_composition}}

@@

\paragraph*{Binding predicates.}
Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
The binding predicate supplied by ioa++ is the ioa::binding\_count function which returns the number of other actions bound to the specified action.
The value returned by ioa::binding\_count is zero for internal actions, zero or one for input actions, and non-negative for output actions.
Most often, ioa::binding\_count is used in the precondition of output actions to ensure that the value generated by the output action will be received if produced.
\ifjournal
This type of usage can be seen in the send action of the asynch\_lcr\_automaton:
\begin{lstlisting}
bool send_precondition () const {
  return !m_send.empty () &&
    ioa::binding_count (&asynch_lcr_automaton::send)
      != 0;
}
\end{lstlisting}
\fi

\subsection{Managing Dynamic Constellations\label{constellations}}

In this section, we introduce techniques and principles for managing dynamic constellations of automata and bindings that go beyond the basic versions presented in the previous discussion.
This list is not exhaustive but rather serves as a distillation of our experiences so far working with ioa++.

\paragraph*{System action events.}
Determining the outcome of a system action event (e.g., that a child automaton was created) is necessary when there are constraints on the order in which system actions must occur.
For example, the Binding Manager mentioned in Section~\ref{system_action_section} waits until the automata that supply the output action and input action are created, before attempting to bind.
Similarly, an automaton might compose with an automaton that it does not manage, and therefore may expect the automaton to exist before attempting to bind.
To enable such behavior, user automata can \emph{observe} ioa::automaton\_manager objects and ioa::binding\_manager objects to receive events generated upon the completion of a system action.

\paragraph*{Error reporting.}
The event semantics of I/O automata provide a natural mechanism for reporting errors.
An automaton that encounters an error can use an output action to signal the error.
The automaton can continue processing if the error is transient or stop if the error is permanent.
In either case, the error is propagated to the automaton that is using the automaton that encountered an error.
Through this technique, errors can be propagated up the hierarchy and resolved at an appropriate level.
The two techniques for handling an error are either to reset or to replace the portion of the constellation that encountered the error.

\paragraph*{Handling errors by resetting.}
An automaton that can be reset has one or more output actions for reporting errors and one or more input actions for resetting.
Automaton designers should strive to write automata that are capable of being reset as this is the simplest error handling strategy.

\paragraph*{Handling errors by replacement.}
The non-deterministic and asynchronous nature of system actions complicates replacing an automaton.
To replace a child automaton, the parent automaton must destroy the child (thus dissolving all bindings with the child), create a new child, and bind to the new child.
Recall that system actions are asynchronous and can be executed in any order.
Thus, the user might request the actions in one order, e.g., destroy, create, bind, but the system may execute them in a different order, e.g., create, bind, destroy.
If the sequence executed by the system automaton is (create, bind, destroy) then the bind will fail because the user automaton's actions are still bound to the old child.
Note that executing the bind action before the create will also fail but this case is prevented by the ioa::binding\_manager class.
Thus, a user automaton must receive notification that the first child has been destroyed before (creating and) binding to the second child.
A general solution is to observe the automaton and wait for it to be destroyed.

>>>>

In this section, we show how I/O automata can be used to construct component based systems using ioa++.
We then discuss the mechanisms available for dynamic composition and techniques for managing dynamic constellations of automata.
As was mentioned in Section~\ref{design}, all user automata inherit from the common base class (ioa::automaton) which implements the system action interface.
\ifjournal
The following code declares the asynch\_lcr\_automaton:
\begin{lstlisting}
template <typename UID>
class asynch_lcr_automaton :
  public ioa::automaton
{
  ...
};
\end{lstlisting}
The asynch\_lcr\_automaton is declared as a C++ template where the UID template parameter must model the \emph{unique identifier} concept\footnote{The term ``concept'' is used here in the sense of generic programming in C++, i.e., a collection of type requirements imposed by a C++ template.}.
The generic programming facilities of the C++ language aid in the development of general purpose reusable components.

@@

\paragraph*{Declaring/defining actions.}
Actions are member variables that contain methods for preconditions, effects, and scheduling along with \emph{traits} that indicate the action's type (input, output, internal), value status, value type, parameter status, and parameter type.
These traits are used to guide the execution of the methods by the dispatcher and to supply values and parameters as necessary.
The traits are also used to check that input actions are bound to output actions with the same value status and type.
Bindings are checked for type compatibility at compile time.
Local actions are declared using three functions corresponding respectively to precondition, effect, and scheduling.
\ifjournal
For example, the send action of the asynch\_lcr\_automaton is declared:
\begin{lstlisting}
private:
  bool send_precondition () const { ... }
  UID send_effect () { ... }
  void send_schedule () const { ... }
public:
  V_UP_OUTPUT (asynch_lcr_automaton, send, UID);
\end{lstlisting}
\fi
Input actions are declared similarly except for the absence of a precondition.
The precondition and scheduling functions are not allowed to change the state of the automaton and are consequently declared const.
The macros for actions that are subject to binding by other automata are declared public.

\paragraph*{Scheduling.}
Automata submit actions to the scheduler using the ioa::schedule function.
The ioa::schedule function can be called at any time.
However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
A common idiom is for each scheduling function in an automaton to call a common member function that schedules all enabled local actions to avoid replicating scheduling logic.
The common scheduling function consists of calls to ioa::schedule guarded by preconditions as described in Section~\ref{design}.

@@

@@

\paragraph*{Binding predicates.}
Recall that automata require the ability to inspect the current set of bindings.
The binding predicate supplied by ioa++, ioa::binding\_count, returns the number of other actions bound to the specified action.
The value returned by ioa::binding\_count is zero for internal actions, zero or one for input actions, and non-negative for output actions.
Most often, ioa::binding\_count is used in the precondition of output actions to ensure that the value generated by the output action will be received if produced.
\ifjournal
This type of usage can be seen in the send action of the asynch\_lcr\_automaton:
\begin{lstlisting}
bool send_precondition () const {
  return !m_send.empty () &&
    ioa::binding_count (&asynch_lcr_automaton::send)
      != 0;
}
\end{lstlisting}
\fi

\paragraph*{Managing Dynamic Constellations}
Determining the outcome of a system action event is necessary when there are constraints on the order in which system actions must occur.
For example, the Binding Manager mentioned in Section~\ref{design} waits until the automata that supply the output action and input action are created, before attempting to bind.
Similarly, an automaton might compose with an automaton that it does not manage, and therefore may expect the automaton to exist before attempting to bind.
To enable such behavior, user automata can \emph{observe} ioa::automaton\_manager objects and ioa::binding\_manager objects to receive events generated upon the completion of a system action.

The event semantics of I/O automata provide a natural mechanism for reporting errors.
An automaton that encounters an error can use an output action to signal the error.
The automaton can continue processing if the error is transient or stop if the error is permanent.
In either case, the error is propagated to the automaton that is using the automaton that encountered an error.
Through this technique, errors can be propagated up the hierarchy and resolved at an appropriate level.
The two techniques for handling an error are either to reset or to replace the portion of the composed automata that encountered the error.
An automaton that can be reset has one or more output actions for reporting errors and one or more input actions for resetting.
To replace a child automaton, the parent automaton must destroy the child (thus dissolving all bindings with the child), create a new child, and bind to the new child.
Recall that system actions are asynchronous and can be executed in any order.
Thus, the user might request the actions in one order, e.g., destroy, create, bind, but the system may execute them in a different order, e.g., create, bind, destroy.
If the sequence executed by the system automaton is (create, bind, destroy) then the bind will fail because the user automaton's actions are still bound to the old child.
Note that executing the bind action before the create will also fail but this case is prevented by the ioa::binding\_manager class.
Thus, a user automaton must receive notification that the first child has been destroyed before (creating and) binding to the second child.
A general solution is to observe the automaton and wait for it to be destroyed.

### 2012-02-24 papers/paper2/related_work.tex
Down to 17 pages.

Our development of ioa++ grew out of the challenges faced when trying to develop distributed systems with threads.
Lee elaborates on the problems faced when using threads for concurrency in~\cite{lee2006problem}.
Given the shift from higher clock frequencies to larger numbers of cores as the driver of hardware performance, these same concerns are relevant to trends that look to deliver performance improvements through increasingly concurrent programs~\cite{sutter2005software}.

Leveraging a formal model of concurrency to make concurrent programming easier is a well established technique.
We consider three exemplars that are most relevant to the discussion in this paper: Communicating Sequential Processes~\cite{hoare1978communicating}, Actors~\cite{agha1986actors}, and I/O automata~\cite{lynch1996distributed}.
Of these, we selected the I/O automata model because of its support for features essential to developing distributed systems.
%% Temporal logic proof techniques~\cite{manna1992temporal},~\cite{lamport1978time} also can be used to reason about I/O automata, towards developing \emph{correct} systems.

@@

To our knowledge, the Spectrum Simulation System~\cite{goldman1990distributed} and the IOA language~\cite{garland2003ioa} are the only existing approaches that allow one to execute a system expressed as a collection of I/O automata.
Spectrum focuses on algorithm development through simulation.
IOA focuses on formal software development including both simulation and verification.

The ioa++ framework can be compared directly to the IOA language due to the availability of a compiler~\cite{tsai2002code},~\cite{tauber2004verifiable},~\cite{tauber2004compiling} that has been used to implement a number of protocols~\cite{georgiou2009automated}.
\cite{georgiou2009automated} lists six challenges when developing the compiler: ``Program structuring,'' ``IOA Programs and external services,'' ``Modeling procedure calls,'' ``Composing automata,'' ``Nondeterminism'', and ``Implementing datatypes.''
Programs written in IOA are compiled into Java and executed on hosts communicating using the MPI library.
IOA requires programs to be presented in a ``node-channel'' form that allows them to be composed with built in ``mediators'' that implement external services such as MPI.
IOA lacks support for procedure calls, which complicates modeling and interfacing with external services, especially when a procedure call may block.
The IOA compiler composes automata statically using a ``composer'' resulting in a ``node automaton.''
The next action to be executed is selected by a function that resolves the non-determinism in the schedule.

IOA and ioa++ represent two fundamentally different approaches to programming directly with I/O automata.
The development of a new language was necessary for IOA due to its focus on formal verification.
% On the other hand ioa++ uses the existing C++ language and compilers.
IOA encapsulates external services using built in ``mediators'' while ioa++ exposes operating system services via file descriptors, relying on the programmer to ensure non-blocking operation.
IOA focuses on static composition to produce a single node program while ioa++ focuses on dynamic composition and allows concurrent execution.
Both IOA and ioa++ require the programmer to add code to schedule actions, which may be a common source of errors.
In IOA, the programmer provides a deterministic scheduling policy.
Scheduling in ioa++, however, is non-deterministic according to the global policy encoded by the controller.

>>>>

Our development of ioa++ grew out of the challenges faced when trying to develop distributed systems with threads.
Lee elaborates on the problems faced when using threads for concurrency in~\cite{lee2006problem}.
Given the shift from higher clock frequencies to larger numbers of cores as the main driver of hardware performance, these same concerns are relevant to trends that look to deliver performance improvements through increasingly concurrent programs~\cite{sutter2005software}.

Formal models of concurrency are also relevant to the discussion in this paper, e.g., Communicating Sequential Processes~\cite{hoare1978communicating}, Actors~\cite{agha1986actors}, and I/O automata~\cite{lynch1996distributed}.
Of these, we selected the I/O automata model because it offers a straightforward representation of reactive systems.
%% Temporal logic proof techniques~\cite{manna1992temporal},~\cite{lamport1978time} also can be used to reason about I/O automata, towards developing \emph{correct} systems.

@@

To our knowledge, the Spectrum Simulation System~\cite{goldman1990distributed} and the IOA language~\cite{garland2003ioa} are the only existing approaches that allow one to execute a system expressed as a collection of I/O automata.
Spectrum focuses on algorithm development through simulation while the focus of ioa++ is to support the direct implementation of real systems.
IOA and ioa++ represent two fundamentally different approaches to programming directly with I/O automata.
The development of a new language was necessary for IOA due to its focus on formal verification.
% On the other hand ioa++ uses the existing C++ language and compilers.
IOA encapsulates external services using built in ``mediators'' while ioa++ exposes operating system services via file descriptors, relying on the programmer to ensure non-blocking operation.
IOA focuses on static composition to produce a single node program while ioa++ focuses on dynamic composition and allows concurrent execution.
Both IOA and ioa++ require the programmer to add code to schedule actions.
In IOA, the programmer provides a deterministic scheduling policy.
Scheduling in ioa++, however, is non-deterministic according to the global policy encoded by the controller.
